Enkephalin-mediated modulation of basal somatic sensitivity by regulatory T cells in mice

Version 1

Abstract

Abstract

CD4+CD25+Foxp3+ regulatory T cells (Treg) have been implicated in pain modulation in various inflammatory conditions. However, the mechanisms by which Treg hamper pain are still unclear. From a meta-analysis of 11 available transcriptomes, we show that the proenkephalin gene (Penk) which encodes the precursor of analgesic opioid peptides, is among the top 10 genes enriched in murine Treg relative to conventional T cells (Tconv). We then show that Penk expression in Treg is under the control of TNFR signaling and the transcription factor BATF. Using mice in which Penk mRNA expression can be tracked with a fluorescent reporter, we also show that Penk expression is restricted to Treg and activated Tconv in non-inflammatory conditions in all examined organs and tissues. Furthermore, inducible ablation of Penk in Treg leads to heat hyperalgesia for both male and female mice. Overall, our results indicate that TNFR signaling and BATF regulation of Penk in Treg might play a key role at modulating basal somatic sensitivity in mice.

Introduction

Regulatory T-cells (Treg), characterized by the expression of the alpha chain of the interleukin-2 receptor CD25 and the transcription factor Foxp3 1, are known to be key players in immuno-regulation in both humans and mice; too few may lead to autoimmune diseases whereas too many may prevent an efficient immune response to cancer2,3. Over the last few years, several new functions of Treg beyond immunoregulation have been identified in tissue regeneration or local regulation of metabolism4–6 for instance. In addition, there are accumulating evidences of a cross-talk between Treg and the nervous system. These includes promotion of oligodendrocyte differentiation or inhibition of neuroinflammation facilitating CNS repair process after brain injuries and preventing cognitive decline7–10. Furthermore, Treg have been involved in the regulation of pain in various model of nerve injury in rats and mice, such as in autoimmune neuritis or chronic constriction of the sciatic nerve11–13. Depletion of Treg have been associated with enhanced pain sensitivity whereas increased Treg number or activity limit pain hypersensitivity14. Although the current view is that Treg control pain through their immuno-suppressive functions 15, whether Treg might regulate pain at steady state is currently unknown. Starting from our previous observation that the expression of the proenkephalin gene (Penk), encoding for analgesic enkephalins, is highly enriched in Treg (Aubert et al., 2019 biorXiv, https://doi.org/10.1101/638072), we characterize the expression of Penk in the immune system and investigate the role of Treg on pain in vivo at steady state using advanced genetically-engineered mouse models. Our results uncover a previously unknown mechanism by which Treg modulate basal somatic sensitivity.

Results and Discussion

A meta-signature of murine Treg and Penk expression in lymphoid and non-lymphoid organs

Using 11 available transcriptomes retrieved from the GEO website, we generated a Treg molecular signature at steady state in lymphoid tissues, improving the robustness of the analysis by several orders of magnitude relative to individual studies. The 25 most differentially expressed genes are depicted in Figure 1A and the complete list is shown in Table S1. As expected from the sorting strategies used to isolate Treg (Table 1 in the Material and Methods section), we observe that Il2ra and Foxp3 are the most differentially expressed genes in Treg compared to Tconv, followed by the well-known Treg markers CTLA-4, Itgae (CD103), Ikzf2 (Helios) and Klrg1. However, several unexpected genes are present in that signature, including the proenkephalin gene Penk, coding for opioid peptides endowed with analgesic properties. Using the Immuno-Navigator database16, a batch-corrected collection of RNA quantification across numerous studies, samples and cell types, we confirm that Penk is highly correlated to Foxp3 in mice (R=0.871, Figure 1B). Enriched expression of Penk in Treg was previously reported in several specific contexts including TCR-transgenic mice17, UVB-exposure6 or in the brain of mice recovering from stroke8. Using publicly available datasets, we also observe that Penk is enriched in Treg in the thymus, the visceral adipose tissue and in muscles (Figure 1C). Thus, Penk mRNA expression in Treg is consubstantial of their generation and is independent of their tissue localization at steady state.

A meta signature of murine Treg and Penk mRNA expression in lymphoid and non lymphoid organs.

A) Top 25 genes enriched in Treg compared to Tconv from lymphoid tissue from at least 10 of the 11 datasets analyzed ranked by fold change (LogFC) of mean expression relative to Tconv. The Penk gene is highlighted in red. B) Correlation of Penk and Foxp3 mRNA expression in all the cell types listed in the legend. The Pearson correlation coefficient is indicated. C) Expression of Penk mRNA in Treg (blue) and Tconv (red) isolated from thymus, visceral adipose tissue (VAT) and muscle. The source of the data is indicated below each graph.

Penk mRNA expression is regulated by TNFR signaling and the BATF transcription factor in Treg

To explore possible mechanisms that could explain the enrichment of Penk mRNA in Treg cells, we examined the Immuno-Navigator dataset to extract the genes that are most correlated with Penk and with each other (Figure 2A). Among these genes, we observe several members of the TNFR superfamily (Tnfrsf1b (TNFR2), Tnfrsf4 (OX40), Tnfrsf9 (4-1BB), Tnfrsf18 (GITR)) as well as the transcription factor Batf. Furthermore, Batf expression is highly correlated with Penk in Treg (R=0.599, Figure 2B). Thus, TNFR and BATF could be involved in the regulation of Penk in Treg. To examine this possibility, we reanalyzed our previously published dataset establishing the transcriptome of Treg stimulated with anti-CD3, anti-CD28 antibodies and TNFR agonists in vitro 18. We observe that addition of TNFR2, OX40 or 4-1BB agonists increases Penk expression at 36h post-stimulation relative to controls (Figure 2C). Interestingly, Batf is also increased by TNFR agonists at an earlier time point (18h), suggesting that TNFR signaling may regulate Penk expression through the BATF transcription factor. Accordingly, a dramatic decrease in Penk expression is observed in Treg lacking BATF 19(Figure 2D). In line with our hypothesis, BATF is a crucial partner of AP-1 transcription factor that has been shown to regulate Penk in the hippocampus20.

Penk mRNA expression is regulated by TNFR signaling and the BATF transcription factor in Treg.

A) A network of the genes most correlated to Penk in Treg is shown. The Pearson correlation values were extracted from the Immuno-Navigator database (taking only Treg for the analysis) and integrated into Cytoscape v3.730. Each node is a gene linked by edges with width proportional to the Pearson correlation. (edge range: 0.538-0.758). B) Illustration of the correlation between expression of Penk and Batf in Treg. Each dot is a sample from the Immuno-Navigator database. The Pearson correlation coefficient is indicated on the figure. C) Penk and Batf mRNA expression after in vitro stimulation of purified Treg with the indicated TNFR agonists prior (0), and at 18 and 36hrs after stimulation. Each dot is a biological replicate from a single experiment. D) GEO2R analysis of the GSE89656 dataset for Penk mRNA variations between wild type control Treg (WT) and BATF-KO Treg.

Penk is prominently expressed by Treg at steady state

Based on these findings, we further investigated the expression of Penk mRNA in various immune cell types across multiple tissues, as defined by a combination of lineage markers. (Figure S1A). To do so, we crossed a transgenic mouse model expressing Tamoxifen (TMX)-inducible Cre recombinase under the promoter of Penk (PenkCre-ERT2) with the ROSA26TdTomato reporter mice. In these mice, any cell that would expressed Penk at the time of TMX administration would become permanently tagged with the tdTomato reporter a few days later. In order to improve detection of those cells by flow cytometry, we also used an anti-mCherry that cross-react with TdTomato (Figure S1B). In lymphoid tissue, such as in lymph node (LN), Penk expression is mainly restricted to Treg and to a small subset of activated CD4+ T-cells (Figure 3A). Similar results are obtained in non-lymphoid tissues, such as in the colon, although Penk expression in activated T cells is further increased relative to LN (Figure 3B). As summarized in Figure 3C, Treg is the immune subset expressing the highest level of Penk in all tissues analyzed. Expression of Penk is also detected in Tconv, but at lower frequencies than in Treg, in agreement with the molecular meta-signature established above (Figure 1). All other cell types show low or undetectable Penk expression. Interestingly, in CD4+ T-cells (Tconv and Treg), Penk expression is higher in the activated CD62Llow CD44high fraction compared to the naive CD62Lhigh CD44low phenotype (Figure S2). Thus, we observed that Penk mRNA is predominantly found in Treg and activated CD4+ T cells in the immune system at steady state. Among these, the highest expression of Penk was detected in Treg cells present in the skin and the colon. However, we cannot exclude the possibility that the distribution of Penk may be broader in an inflammatory context. In line with this, a recent study demonstrated that Penk expression in M2 macrophages is dependent on IL-4 and occurs in response to nerve injury21.

Penk is prominently expressed by Treg at steady state

A-B) (right) UMAP representing all major cell types indicated in the figures determined by flow cytometry from lymph nodes (A) and colon (B) (left). Projection of Penk (blue) on the UMAP of lymph node and colon (right). Subsets were manually gated as depicted in Figure S1A. C) Scatter plot displaying the average population size and percentage of Penk+ cells for the listed cell populations and organs. Population size was calculated as the percentage out of total CD45+ single cells and represented on a log10 scale. (n = 3 mice for the VAT, spleen and bone marrow, n = 6 mice for the other groups; results cumulative of two independent experiments).

Heat hyperalgesia in mice deficient for Penk in Treg

To determine if enkephalins produced by Treg had an effect on pain at steady state, we generated mice deficient for enkephalins in Treg by crossing TMX-inducible Cre recombinase under the control of the Foxp3 promoter (Foxp3Cre-ERT2) with mice transgenic for LoxP sequences flanking exon 2 of Penk (Penkflox). In these mice (hereafter referred to as Lox), any cell expressing Foxp3 at the time of TMX administration would become deficient for exon 2 of Penk a few days later. Since exon 2 is the precursor of Met-Enkephalin, an endogenous opioid that act on thermal pain sensation22, we evaluated the sensibility of these mice to pain induced by heat (Figure 4). As controls, we used mice expressing the Cre recombinase and wild-type (WT) at the locus of LoxP sequences insertion. Mice were treated with TMX and evaluated for heat sensitivity without further additional treatment at ten different time points (4 before and 6 after administration of TMX, 2 to 3 days apart). Under these conditions, a significant trend towards lower latency periods (signing heat hyperalgesia) is observed in the Lox group compared to WT mice. Interestingly, the effect is not apparent until 7 days after TMX administration (Figure S4). Indeed, before TMX administration, WT and Lox female and male mice do not differ in their response to heat (Figure 4A-B). However, after TMX administration, Lox mice develop hyperalgesia compared to control mice, with a 20 percent reduction in their median latency period from D7 onwards (8.1 vs. 6.5 sec, Figure 4C). Interestingly, the effect of Penk deletion in Treg on heat hyperalgesia is sex independent, since it is observed in both female and male (Figure 4D), although the reduction was slightly higher in female (22% reduction in latency period) than in male (16% reduction). This can be explained by the fact that males tend to respond faster (7.9s vs 6.1s) than females (8.4s vs 7.0s), irrespective of the genotype (Figure 4D). Penk expression by CD4+ T-cells has been associated with analgesia in mice models of visceral pain23–25, but the specific contribution of Treg was not investigated. In neuroinflammatory settings, such as constriction of the sciatic nerve, pain control by Treg could be dependent on their immuno-suppressive function 12,26–28. On the contrary, our tests were performed at steady state, ruling out the possibility that hyperalgesia could be due to a defect in the immuno-suppressive function of Treg. They strongly suggest a direct role of enkephalins produced by Treg on nociception, uncovering a new non-immune fundamental function of Tregs on endogenous regulation of basal somatic sensitivity.

Heat hyperalgesia in mice deficient for Penk in Treg.

A) Withdrawal latency of WT and Lox mice before the administration of TMX (baseline). The mean of all 4 measures before administration of TMX for each mouse is shown. Statistical modeling was performed using a non parametric unpaired Mann-Whitney t-test and was deemed as non-significant (ns). B) Withdrawal latency of WT and Lox mice before administration of TMX (baseline) shown separately for female and male mice, as in A. Statistical modeling was performed using two-way ANOVA and the effect of the genotype on the results was deemed non-significant (ns). C) Withdrawal latency of WT and Lox mice post-TMX treatment. The mean of all 4 measures taken from D7 onward is represented for each mouse. Statistical modeling was performed using a non parametric unpaired Mann-Whitney t-test and was deemed significant (p=0,0038). D) Withdrawal latency of WT and Lox mice post-TMX treatment shown separately for female and male mice, as in C. Statistical modeling was performed using two-way ANOVA and the effect of the genotype on the results was deemed significant (p=0,0037). A multiple comparisons test using the Benjamini-Hochberg False Discovery Rate model was used. The adjusted p values were deemed significant for both female and male mice (q=0,043). Results shown in this figure are cumulative of two independent experiments with n = 44 mice (26 WT and 18 Lox).

eLife Assessment

This study presents a valuable finding on a new role of Foxp3+ regulatory T cells in sensory perception, which may have an impact on our understanding of somatosensory perception. The authors identified a previously unappreciated action of enkephalins released by immune cells in the resolution of pain and several upstream signals that can regulate the expression of the proenkephalin gene PENK in Foxp3+ Tregs. However, whereas the generation of transgenic mice with conditional deletion of PENK in Foxp3+ cells and PENK fate-mapping is novel and generates compelling data, they show an incomplete analysis of Tregs in the control and transgenic mice, proper tamoxifen controls nor the role of PENK+ skin T cells to further support their hypothesis. Nonetheless, the study would be of interest to the biologists working in the field of neuroimmunology and inflammation.

Reviewer #1 (Public Review):

Summary:

The authors explore mechanisms through which T-regs attenuate acute pain using a heat sensitivity paradigm. Analysis of available transcriptomic data revealed expression on the proenkephalin (Penk) gene in T-regs. The authors explore the contribution of T-reg Penk in the resolution of heat sensitivity.

Strengths:

Investigating the potential role of T-reg Penk in the resolution of acute pain is a strength.

Weaknesses:

The overall experimental design is superficial and lacks sufficient rigor to draw any meaningful conclusions.

For instance:

1. The were no TAM controls. What is the evidence that TAM does not alter heat-sensitive receptors.

2. There are no controls demonstrating that recombination actually occurred. How do the authors know a single dose of TAM is sufficient?

3. Why was only heat sensitivity assessed? The behavioral tests are inadequate to derive any meaningful conclusions. Further, why wasn't the behavioral data plotted longitudinally

Reviewer #2 (Public Review):

Summary:

The present study addresses the role of enkephalins, which are specifically expressed by regulatory T cells (Treg), in sensory perception in mice. The authors used a combination of transcriptomic databases available online to characterize the molecular signature of Treg. The proenkephalin gene Penk is among the most enriched transcripts, suggesting that Treg plays an analgesic role through the release of endogenous opioids. In addition, in silico analysis suggests that Penk is regulated by the TNFR superfamily; this being experimentally confirmed. Using flow cytometry analysis, the authors then show that Penk is mostly expressed in Treg of the skin and colon, compared to other immune cells. Finally, genetic conditional excision of Penk, selectively in Treg, results in heat hypersensitivity, as assessed by behavior analysis.

Strengths:

The manuscript is clear and reveals a previously unappreciated role of enkephalins, as released by immune cells, in sensory perception. The rationale in this manuscript is easy to follow, and conclusions are well supported by data.

Weaknesses:

The sensory deficit of Penk cKO appears to be quite limited compared to control littermates.

Reviewer #3 (Public Review):

Summary:

Aubert et al investigated the role of PENK in regulatory T cells. Through the mining of publicly available transcriptome data, the authors confirmed that PENK expression is selectively enriched in regulatory but not conventional T cells. Further data mining suggested that OX40, 4-1BB as well as BATF, can regulate PENK expression in Tregs. The authors generated fate-mapping mice to confirm selective PENK expression in Tregs and activated effector T cells in the colon and spleen. Interestingly, transgenic mice with conditional deletion of PENK in Tregs resulted in hypersensitivity to heat, which the authors attributed to heat hyperalgesia.

Strengths:

The generation of transgenic mice with conditional deletion of PENK in foxp3 and PENK fate-mapping is novel and can potentially yield significant findings. The identification of upstream signals that regulate PENK is interesting but unlikely to be the main reason why PENK is predominantly expressed in Tregs as both BATF and TNFR are expressed in effector T cells.

Weaknesses:

There is a lack of direct evidence and detailed analysis of Tregs in the control and transgenic mice to support the authors' hypothesis. PENK was previously reported to be expressed in skin Tregs and play a significant role in regulating skin homeostasis: this should be considered as an alternative mechanism that may explain the changed sensitivity to heat observed in the paper.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The authors explore mechanisms through which T-regs attenuate acute pain using a heat sensitivity paradigm. Analysis of available transcriptomic data revealed expression on the proenkephalin (Penk) gene in T-regs. The authors explore the contribution of T-reg Penk in the resolution of heat sensitivity.

Strengths:

Investigating the potential role of T-reg Penk in the resolution of acute pain is a strength.

Weaknesses:

The overall experimental design is superficial and lacks sufficient rigor to draw any meaningful conclusions.

We hope that the reviewer will reconsider this severe criticism after examining the updated manuscript and results.

For instance:

(1) The were no TAM controls. What is the evidence that TAM does not alter heat-sensitive receptors.

the impact of TMX on heat perception is not the object of this study. Nevertheless, it appears that heat-sensitivity in controls WT (blue dots) is slightly diminished after TMX administration (Figure 5A), suggesting that heat-sensitive receptors are moderately altered by TMX per se. This reduction is much more pronounced for LOX mice. Thus, although it is possible that TMX play a marginal role on heat sensitivity by itself, the results show a much more pronounced effect of TMX in LOX than in WT, in favor of a role for Penk Treg in heat sensitivity.

(2) There are no controls demonstrating that recombination actually occurred. How do the authors know a single dose of TAM is sufficient?

these results are now presented in figure S4. A 70% reduction in Penk mRNA is observed in Treg after a single administration of TMX.

(3) Why was only heat sensitivity assessed? The behavioral tests are inadequate to derive any meaningful conclusions. Further, why wasn't the behavioral data plotted longitudinally

The longitudinal data are presented in figure S5A. New behavioral tests have been performed and the results are now shown in figure S5E-H. Importantly, heat sensitivity was observed in two independent laboratory with two different tests.

Reviewer #2 (Public Review):

Summary:

The present study addresses the role of enkephalins, which are specifically expressed by regulatory T cells (Treg), in sensory perception in mice. The authors used a combination of transcriptomic databases available online to characterize the molecular signature of Treg. The proenkephalin gene Penk is among the most enriched transcripts, suggesting that Treg plays an analgesic role through the release of endogenous opioids. In addition, in silico analysis suggests that Penk is regulated by the TNFR superfamily; this being experimentally confirmed. Using flow cytometry analysis, the authors then show that Penk is mostly expressed in Treg of the skin and colon, compared to other immune cells. Finally, genetic conditional excision of Penk, selectively in Treg, results in heat hypersensitivity, as assessed by behavior analysis.

Strengths:

The manuscript is clear and reveals a previously unappreciated role of enkephalins, as released by immune cells, in sensory perception. The rationale in this manuscript is easy to follow, and conclusions are well supported by data.

Weaknesses:

The sensory deficit of Penk cKO appears to be quite limited compared to control littermates.

Reviewer #3 (Public Review):

Summary:

Aubert et al investigated the role of PENK in regulatory T cells. Through the mining of publicly available transcriptome data, the authors confirmed that PENK expression is selectively enriched in regulatory but not conventional T cells. Further data mining suggested that OX40, 4-1BB as well as BATF, can regulate PENK expression in Tregs. The authors generated fate-mapping mice to confirm selective PENK expression in Tregs and activated effector T cells in the colon and spleen. Interestingly, transgenic mice with conditional deletion of PENK in Tregs resulted in hypersensitivity to heat, which the authors attributed to heat hyperalgesia.

Strengths:

The generation of transgenic mice with conditional deletion of PENK in foxp3 and PENK fate-mapping is novel and can potentially yield significant findings. The identification of upstream signals that regulate PENK is interesting but unlikely to be the main reason why PENK is predominantly expressed in Tregs as both BATF and TNFR are expressed in effector T cells.

Weaknesses:

There is a lack of direct evidence and detailed analysis of Tregs in the control and transgenic mice to support the authors' hypothesis. PENK was previously reported to be expressed in skin Tregs and play a significant role in regulating skin homeostasis: this should be considered as an alternative mechanism that may explain the changed sensitivity to heat observed in the paper.

We now provide a detailed analysis of Treg with or without Penk, from their immunosuppressive functions to their colocalization with sensory neurons in the skin, supporting their function as natural analgesics. The alternate hypothesis relative to skin homeostasis is now clearly presented and discussed.

Recommendations for the authors):

Reviewer #2 (Recommendations For The Authors):

Most of my comments should be addressable in a revised manuscript but will require additional analysis.

Major:

- According to flow cytometry analysis, Penk is expressed mostly in Treg of the skin and colon. What may account for such restricted expression? Where could Treg-released enkephalins act?

We now rephrased the paper to emphasize the known role of Batf in tissue Treg differentiation. We believe the Batf dependency of Penk expression is the reason why tissue Treg are more enriched in Penk than Treg from lymphoid organs. This is now clearly discussed.

We also provide a new figure (Figure S1) that shows that binding of Batf and co factors AP1 and IRF4 were reported to bind to Penk regulatory regions. Altogether, the role of Batf in tissue Treg differentiation would explain why tissue Treg such as colon and skin are particularly enriched in Penk. This is now clearly stated in the revised manuscript.

As to know where Treg-released enkephalins act, we performed immunostainings in the skin and observed that Treg could colocalize with sensory neurons (shown in a new figure 5, panel D). This observation raise the hypothesis that Treg-released enkephalins could act on sensory neurons locally.

- Which mechanism can underlie heat hypersensitivity in Penk cKO mice? Which sensory neurons are involved? Are other sensory modalities affected, such as mechanical sensitivity?

As stated above, we show that Treg can be in close contact with thermal sensors neurons producing CGRP. These data are shown in figure 5D. We have also tested may other nociceptive stimulus (innocuous and noxious) and did not detect significant differences. These data are presented as a supplementary figure S5. Whether enkephalins produced by Treg can change the stimulation threshold of various nervous fibers is currently performed by electrophysiology.

- No control is provided to ensure that Penk is selectively excised in Treg cells in cKO mice.

We have performed additional experiments with fluorescent probes to document Penk mRNA expression in cKO mice. The results on the specific expression of Penk mRNA in various subsets post-TMX are shown in a supplementary figure S4.

- The authors acknowledge that Penk from Treg was previously studied in an animal model of inflammatory pain. However, which role these endogenous opioids play is unclear, especially since authors discovered that enkephalins are likely continuously released at steady states. This is not enough discussed in the narrative, which surprisingly does not separate the results from the discussion.

The results and discussion are now separated in two sections.

Minors:

- Replace "Fox3 1" with "Fox31" (line 31), "functions 15" with "functions15" (line 43), "BATF 19" with "BATF19" (line 85).

- Text mentions Figure S4 (line 125), which is most likely S3.

Reviewer #3 (Recommendations For The Authors):

Given the most significant finding of this paper is based on the heat-induced pain model, there is surprisingly little analysis of Tregs in this context. The authors analyzed spleen and colon Tregs at steady state, it is unclear whether any of these Tregs are involved in pain sensitivity directly. Skin Tregs or other relevant Tregs to this model should be analyzed in control and Lox mice. This is particularly relevant as PENK expression was previously reported in skin Tregs and plays a significant role in skin homeostasis (Yamazaki et al 2020 PNAS). Does PENK conditional deletion alter Treg frequencies, numbers, and immune suppressive function? Not even spleen or colon Treg were analyzed comparing control and lox mice.

We now provide evidences showing unaltered immunosuppressive functions of Treg in the absence of Penk (Figure 4), and more importantly unaffected proportions of skin Treg in mice lacking Penk in Treg, at the very site of heat stimulation (Figure 5B-C). We also observed unaffected representation of Treg in the spleen and lymph nodes, but we do not feel that these data are necessary to interpret the results.

Given the role of PENK in skin Tregs, could the observed effect in Figure 4 be due to altered skin homeostasis rather than sensitivity to pain?

The reviewer is referring to a paper where Penk in skin Treg play a role on UV-damaged keratinocytes in vivo (Shime et al., 2020, PNAS). To our knowledge, a role for Penk produced by skin Treg on keratinocytes homeostasis at the steady state is currently unknown. Nevertheless, this hypothesis is now clearly stated and discussed in the manuscript.

The authors stated that only after 7 days post tamoxifen treatment was heat hyperalgesia observed: deletion of PENK in Treg but not Tconv should be confirmed: is deletion only complete after 7 days or is the effect observed due to indirect effects of altered "normal" Treg function?

We have performed a kinetics to document Penk deletion at D3, D7 and 30 post-TMX. Results show a specific deletion of Penk in Treg at all time points so we combined all the time points for the representation of the results (Figure S4). As for the indirect effects of “altered” normal function, we now provide the reader with a new figure (Figure 4), showing that Penk deficient Treg are not impaired in their suppressive function in vitro and in vivo.

Actual protein/peptide production of enkephalins by Tregs should be confirmed. It is also unclear which peptide(s) can be secreted and presumably responsible for the changes in heat sensitivity.

This is a very interesting question that we addressed with a MENK ELISA but without success at reproducing the results. An ongoing project will use mass spectrometry to fully characterize the peptides produced by Treg and activated Tconv.

The analysis of PENK regulation by Tregs is interesting despite them being entirely based on data mining. BATF is a pioneering factor expressed by all activated effector T cells. While the connection between BATF and PENK may explain why the authors observed PENK expression chiefly in activated effectors and Tregs, BATF cannot be the reason why PENK is "predominantly" expressed by Tregs. Similarly, 4-1BB and OX40 can be induced on effector T cells. Is PENK under the control of Foxp3? There are lots of publically available datasets on Foxp3/IL-2 dependent Treg signatures through which this can be addressed.

We now provide a supplementary figure (Figure S1), showing a compilation of ChIP Seq studies for various transcription factors in various T cell subsets. We provide the reader with a list of all the TF that have been reported to bind in the regulatory regions of Penk. In agreement with our hypothesis, BATF, FOXP3, IRF4 and several others are present in that list. Further work is needed to decipher the exact contribution of each of those TF to the regulation of Penk in Treg vs activated Tconv that is beyond the scope of this report.

Version 2

Abstract

Abstract

CD4+CD25+Foxp3+ regulatory T cells (Treg) have been implicated in pain modulation in various inflammatory conditions. However, whether Treg cells hamper pain at steady state and by which mechanism is still unclear. From a meta-analysis of the transcriptomes of murine Treg and conventional T cells (Tconv), we observe that the proenkephalin gene (Penk), encoding the precursor of analgesic opioid peptides, ranks among the top 25 genes most enriched in Treg cells. We then present various evidence suggesting that Penk is regulated in part by members of the TNF receptor family and the transcription factor Batf. Using mice in which the promoter activity of Penk can be tracked with a fluorescent reporter, we also show that Penk expression is mostly detected in Treg and activated Tconv in non-inflammatory conditions in the colon and skin. Functionally, Treg cells proficient or deficient for Penk suppress equally well the proliferation of effector T cells in vitro and autoimmune colitis in vivo. In contrast, inducible ablation of Penk in Treg leads to heat hyperalgesia in both male and female mice. Overall, our results indicate that Treg might play a key role at modulating basal somatic sensitivity in mice through the production of analgesic opioid peptides.

Introduction

Regulatory T-cells (Treg), characterized by the expression of the alpha chain of the interleukin-2 receptor CD25 and the transcription factor Foxp3 (Hatzioannou et al., 2021), are known to be key players in immuno-regulation in both humans and mice; too few may lead to autoimmune diseases whereas too many may prevent an efficient immune response to cancer (Kim et al., 2007; Nishikawa and Sakaguchi, 2010).

Over the last few years, several new functions of Treg beyond immunoregulation have been identified in tissue regeneration or local regulation of metabolism (Xiao et al., 2022; Meng et al., 2023; Shime et al., 2020) for instance. In addition, there is accumulating evidence of a cross-talk between Treg and the nervous system. This includes promotion of oligodendrocyte differentiation or inhibition of neuroinflammation facilitating CNS repair process after brain injuries and preventing cognitive decline (Dombrowski et al., 2017; Huang et al., 2020; Ito et al., 2019; Lemaitre et al., 2023).

Furthermore, Treg have been involved in the regulation of pain in various models of nerve injury in rats and mice, such as in autoimmune neuritis or chronic constriction of the sciatic nerve (Austin et al., 2012; Duffy et al., 2019; Kuhn et al., 2021). Depletion of Treg has been associated with enhanced pain sensitivity whereas increased Treg number or activity limit pain hypersensitivity (Lees et al., 2015). Although the current view is that Treg controls pain through their immuno-suppressive functions (Bethea and Fischer, 2021), whether Treg might regulate pain at steady state is currently unknown. Our results uncover a previously unknown function of Treg that modulates basal somatic sensitivity through the production of analgesic peptides derived from the proenkephalin Penk gene.

Results

A meta-signature of murine Treg

Using 11 available transcriptomes retrieved from the GEO website, we generated a Treg molecular signature at steady state in lymphoid tissues, significantly enhancing the robustness of our analysis compared to individual studies. The 25 most differentially expressed genes are depicted in Figure 1A, with the complete list provided in Table S1. As expected from the sorting strategies used to isolate Treg (detailed in Table 1 of the Materials and Methods section), Il2ra and Foxp3 emerged as the most differentially expressed genes in Treg compared to Tconv. Several well-known Treg markers such as CTLA-4, Itgae (CD103), Ikzf2 (Helios), Tnfrsf4 (OX40) and Tnfrsf9 (4-1BB) are also present on this list. Some genes, such as Gpr83 and Rgs1, have been associated with Treg functions (Flynn et al., 2023; Lu et al., 2007) though their exact roles remain to be fully elucidated. Interestingly, the majority of genes on this list, including Fam129a, Coro2a, Osbpl3, and Penk, have unknown functions in Treg.

A meta signature of murine Treg and Penk mRNA expression in lymphoid and non-lymphoid organs.

A) Top 25 genes enriched in Treg compared to Tconv from lymphoid tissue from at least 10 of the 11 datasets analyzed ranked by fold change (LogFC) of mean expression relative to Tconv. The Penk gene is highlighted in red. B) Correlation of Penk and Foxp3 expression in all the cell types listed in the legend according to the Immuno-navigator dataset. The Pearson correlation coefficient is indicated. Each dot is a sample, color coded as a subset according to the legend. C) Expression of Penk in Treg (blue) and Tconv (red) isolated from thymus, visceral adipose tissue (VAT) and muscle. The source of the data is indicated below each graph.

Penk expression in Treg of lymphoid and non-lymphoid organs and tissues

Using the Immuno-Navigator database (Vandenbon et al., 2016), which provides a batch-corrected collection of RNA quantification across numerous studies, samples and cell types, we confirm that Penk is highly correlated to Foxp3 in lymphoid organs of mice (r=0.871, Figure 1B). Notably, Treg samples exhibited the highest level of both Foxp3 and Penk. While it has been previously reported that Th2 and Th1 cells can express Penk (Boué et al., 2012), they do so to a lesser extent than Treg (Figure 1B).

Additionally, using publicly available datasets comparing Treg and Tconv, we observe that Penk is enriched in the thymus, where Treg are generated, and is also present in peripheral tissues such as visceral adipose fat and muscles (Figure 1C). Therefore, the enrichment of Penk mRNA in Treg is intrinsic to their generation and is independent of their tissue localization at steady state.

Penk mRNA expression is regulated by TNFR signaling and the BATF transcription factor

To explore possible mechanisms explaining the enrichment of Penk mRNA in Treg cells, we examined the Immuno-Navigator dataset to identify genes most correlated with Penk and with each other in Treg samples. We represent these correlations as a network where each node is a gene and each edge is a correlation above a certain threshold (Figure 2A). Penk expression is directly correlated to the expression of 5 genes: the TNF receptor family members, Tnfrsf4 (OX40), Tnfrsf9 (4-1BB), Tnfrsf18 (GITR), the transcription factor Batf and the short chain dehydrogenase/reductase family 39U member 1 Sdr39u1. As an illustration, we show the correlation between Batf and Penk expression in Treg samples (r=0.599, Figure 2B).

Penk expression is regulated by TNFR signaling and the BATF transcription factor.

A) A network of the genes most correlated to Penk in Treg is shown. The Pearson correlation values were extracted from the Immuno-Navigator database (selecting only Treg in the analysis) and integrated into Cytoscape v3.7 (Shannon et al., 2003). Each node is a gene linked by edges with width proportional to the Pearson correlation. (edge range: 0.538-0.758). B) Illustration of the correlation between expression of Penk and Batf in Treg. Each dot is a sample from the Immuno-Navigator database. The Pearson correlation coefficient is indicated on the figure. C) Penk and Batf mRNA expression after in vitro stimulation of purified Treg with the indicated TNFR agonists prior (0), and at 18 and 36 hrs after stimulation. Each dot is a biological replicate from a single experiment. D) GEO2R analysis of the GSE89656 dataset between wild type control Treg (WT) and BATF-KO Treg.

Furthermore, Penk is indirectly correlated with Tnfrsf1b (TNFR2), Il2ra (CD25) and Cish, a negative regulator of cytokine signaling. These strong correlations between several TNFR members, the transcription factor Batf, and Penk suggest a possible regulatory pathway. To explore this possibility, we reanalyzed our previously published dataset on the transcriptome of Treg stimulated with anti-CD3, anti-CD28 antibodies and TNFR agonists in vitro (Lubrano di Ricco et al., 2020). We observe that addition of TNFR2, OX40 or 4-1BB agonists increases Penk expression at 36 hours post-stimulation relative to controls (Figure 2C). Interestingly, Batf is also increased with TNFR agonists but at an earlier time point (18 hours). Consistent with our hypothesis, a dramatic decrease in Penk expression is observed in Treg lacking Batf (Hayatsu et al., 2017) (Figure 2D).

Furthermore, analysis of the UniBind database (Puig et al., 2021) revealed that the transcription factors Batf, Irf-4, Jun and Fosl2 (AP-1 members), RelA (NF-kB signaling), and the master Treg regulator Foxp3 have all been shown by ChIP-Seq to bind to the promoter/enhancer regions of Penk in various T-cell subsets (Figure S1).

Penk is predominantly expressed by Treg at steady state

Among other roles, Batf has been linked to tissue Treg differentiation in mice (Burton et al., 2023; Delacher et al., 2020; Hayatsu et al., 2017). Thus, we hypothesize that tissue Treg might be further enriched in Penk relative to lymphoid organs. To track Penk-expressing cells in vivo, we crossed a transgenic mouse model expressing Tamoxifen (TMX)-inducible Cre recombinase under the promoter of Penk (PenkCre-ERT2) with the ROSA26TdTomato reporter mice. In these mice, any cell that expressed Penk at the time of TMX administration would become permanently tagged with the tdTomato reporter a few days later.

We investigated the expression of Penk mRNA in various immune cell types across multiple tissues by spectral flow cytometry, using a combination of lineage markers (Table S2) and an appropriate gating strategy (Figure S2A). To improve detection of Penk mRNA-expressing cells, we also used an anti-mCherry that cross-reacts with TdTomato (Figure S2B). In a UMAP representation of high-dimensional flow cytometry data, the projection of tdTomato expression (indicating Penk expression) aligns with the clusters of Treg and a small subset of activated CD4+ T-cells in the lymph nodes (LN) (Figure 3A). Compared to LN, Penk expression encompasses entire clusters of Treg and activated T cells in the colon, likely due to the lower proportion of naïve T cells in this tissue (Figure 3B). Interestingly, in CD4+ T-cells (Tconv and Treg), Penk expression is higher in the activated CD62Llow CD44high fraction compared to the naive CD62Lhigh CD44low phenotype (Figure S3). As summarized in Figure 3C, the highest frequency of Penk+ cells is observed in Treg across all analyzed tissues, with the highest frequencies in the colon and the skin. Penk expression is also detected in Tconv of the colon and skin, but at lower frequencies than in Treg. All other cell types show low or undetectable Penk expression.

Penk is predominantly expressed by Treg at steady state.

A-B) (Left) UMAP representing all major cell types indicated in the figures determined by flow cytometry from lymph nodes (A) and colon (B). (Right) Projection of tdTomato expression on the UMAP of lymph nodes and colon. Subsets were manually gated as depicted in Figure S1A. C) Bubble plot displaying the average population size and frequencies of Penk+ cells for the listed cell populations and organs. Population size was calculated as the percentage out of total CD45+ single cells and represented on a log10 scale. (n = 3 mice for the VAT, spleen and bone marrow, n = 6 mice for the other groups; results cumulative of two independent experiments).

Immunosuppressive functions of Treg are unaffected by the lack of enkephalins

To test whether the lack of enkephalins in Treg impacts their suppressive function, we generated mice deficient in enkephalins in the hematopoietic compartment by grafting bone marrow from Penk knock-out mice in immunodeficient RAG2-KO mice. Several months after the graft, Treg were sorted from lymphoid organs and tested in vitro for their ability to suppress the proliferation of conventional T cells. No significant difference in the suppression of Tconv proliferation is observed between Treg proficient or deficient for Penk (Figure 4A). Similarly, the addition of Naloxone, an irreversible blocker of enkephalin receptors, neither abolishes nor enhances the suppressive function of normal Treg cells (Figure 4B). Additionally, both Penk-KO and wild-type Treg isolated from these chimeric mice equally prevented the occurrence of autoimmune colitis induced by the transfer of naive Tconv cells into RAG2-KO mice (Figure 4C).

Immunosuppressive functions of Treg are unaffected by the lack of enkephalins.

A) In vitro suppression of WT Tconv proliferation by WT or Penk KO Tregs. B) In vitro suppression of WT Tconv proliferation by Penk WT Tregs in presence of Naloxone. C) Body weight and D) survival of Rag2-/- mice transferred with Tconv and Treg from Penk-WT or Penk-KO chimeric mice (2:1 ratio), as described in the methods. All data are cumulative of 2 independent experiments.

Overall, these results indicate that enkephalins are not major players in the suppressive functions of Treg cells both in vitro and in vivo.

Heat hyperalgesia in mice deficient for Penk in Treg

To our knowledge, the analgesic function of Penk in Treg has never been reported. To determine if enkephalins produced by Treg affect pain at steady state, we generated mice deficient in Penk by crossing Tamoxifen (TMX)-inducible Cre recombinase under the control of the Foxp3 promoter (Foxp3Cre-ERT2) with mice transgenic for LoxP sequences flanking exon 2 of Penk (Penkflox). In these mice (hereafter referred to as LOX), any cell expressing Foxp3 at the time of TMX administration would become deficient for exon 2 of Penk a few days later, hence lacking enkephalins.

Using flow cytometry, we observe that Penk mRNA expression is reduced by more than half in TMX-treated LOX mice compared to WT mice (Figure S4). Consistent with the Penk-Cre reporter mouse data, Penk mRNA expression is very low in non-Treg (CD4+Foxp3- or CD8+ cells) and does not differ significantly between LOX and WT mice. Thus, TMX treatment specifically reduces Penk mRNA expression in Treg cells of LOX mice.

Since exon 2 is the precursor of Met-Enkephalin, an endogenous opioid that affects thermal pain sensation (Aman et al., 2016), we evaluated the sensitivity of these mice to pain induced by heat (Figure 5). As controls, we used mice expressing the Cre recombinase and wild-type (WT) at the locus of LoxP sequences insertion. Mice were treated with TMX and evaluated for heat sensitivity at 10 different time points (four before and six after administration of TMX, 2 to 3 days apart). Under these conditions, a significant trend towards lower latency periods (indicative of heat hyperalgesia) is observed in the LOX group compared to WT mice (Figure S5A). Interestingly, the effect is not apparent until seven days after TMX administration. Before TMX administration, WT and LOX mice do not differ in their response to heat (Figure 5A). However, after TMX administration, LOX mice develop hyperalgesia compared to WT mice, with a 20 percent reduction in their median latency period from day 7 onwards ((8.1 vs. 6.6 sec.), Figure 5A). The effect of Penk deletion in Treg on heat hyperalgesia is sex-independent, as it is observed in both females and males (Figure S5B-C). Moreover, this thermal hyperalgesia in LOX mice is reproduced with a different test in an independent laboratory where WT and LOX mice were sent for further behavioral tests (Figure S5D). Indeed, tests for innocuous (von Frey, cotton swab tests) and noxious (pin prick test) mechanical sensitivity, as well as light touch and proprioception (sticky tape test), fail to show any significant effect of Treg-specific Penk deficiency (Figure S5E-H). Importantly, TMX administration does not alter the proportions of skin Treg in LOX compared to WT mice (Figure 5B-C), indicating that hyperalgesia does not result from an altered distribution of Treg.

Heat hyperalgesia in mice deficient for Penk in Treg.

A) Withdrawal latency of WT and LOX mice before (baseline) and after administration of Tamoxifen (TMX). Each dot corresponds to the mean latency response (in seconds) of 4 measurements taken before TMX administration (baseline) and 4 measurements taken from day 7 onwards (TMX). Statistical modeling was performed using a non-parametric unpaired Mann-Whitney t-test with multiple corrections. The results shown in this figure are cumulative from two independent experiments with a total of 44 mice (26 WT and 18 LOX). Each dot is a mouse. B) Representative flow cytometry contour plot of CD25 and Foxp3 staining on pad skin CD45+CD3+CD4+ cells from WT or LOX mice 17 days after TMX gavage. C) Quantification of the frequency of Treg (Foxp3+CD25+) among CD4+ T cells in WT and LOX mice pad as shown in (B). The indicated p value was determined by a Mann-Whitney test. Each dot is a mouse from a single experiment. D) Immunofluorescence staining of CGRP neurons (red), Foxp3-GFP cells (green), and DAPI (blue) of footpad skin section of a female LOX mouse (scale bar represents 50µm). The right panel is the magnification of the area indicated on the left panel.

Given this result, we next investigated whether Treg could be localized in contact with nociceptive neurons in the skin. Interestingly, some skin Tregs marked by the GFP reporter molecule in our WT and LOX mice can be observed in close contact with free nerve endings labeled with a Calcitonin Gene-Related Peptide (CGRP)-specific mAb in the pad skin (Figure 5D). Because sensory neurons expressing CGRP are essential for noxious thermal heat, but not mechanical sensitivity (McCoy et al., 2013), this result suggests that Treg producing enkephalins could act locally on nociception.

Discussion

The proenkephalin gene Penk encodes the precursor of opioid peptides with analgesic properties (McLaughlin, 2013). Enriched expression of Penk in Treg has been previously reported in several specific contexts, including TCR-transgenic mice (Zelenika et al., 2002), UVB-exposure (Shime et al., 2020) or in the brains of mice recovering from stroke (Ito et al., 2019) but its role has never been directly addressed. We first explore the possible molecular mechanisms that may explain the preferential expression of Penk observed in tissue Treg. Using data mining and gene correlation analysis, we observed that TNFR and Batf might be involved in Penk regulation. Batf is known to regulate several genes through partnering with AP-1 and Irf-4 (Murphy et al., 2013), and we noted that several ChIP-Seq studies have reported the binding of these transcription factors in regulatory regions of Penk. Thus, our results support the hypothesis that TNFR signaling may regulate Penk expression in murine T cells through cooperation between Batf, AP-1 and/or Irf-4. Supporting this hypothesis, the AP-1 members Fos and Jun are crucial in Penk regulation in the murine hippocampus (Sonnenberg et al., 1989). Additionally, analysis of Penk-Cre reporter mice led us to conclude that Penk mRNA is predominantly found in tissue Treg, further supporting the hypothesis that Batf might be a chief regulator of Penk. Although Penk might be preferentially expressed by activated T cells at steady state (Treg and Tconv), its distribution may be broader in an inflammatory context. Consistent with this, it has been reported that IL-4-treated macrophages are able to reduce neuropathic pain through their ability to produce opioid peptides (Celik et al., 2020; Pannell et al., 2016).

A prior study attributed a function to UVB-exposed Treg-derived Penk in promoting the growth of epidermal keratinocytes in vitro and facilitating wound-healing in vivo (Shime et al., 2020). Consequently, heightened heat sensitivity in the absence of Treg-derived Penk may result from altered keratinocyte homeostasis in vivo. Nevertheless, the impact of Treg-derived Penk on keratinocyte homeostasis in vivo under normal conditions has yet to be conclusively demonstrated.

Penk expression by CD4+ T-cells has been linked to analgesia in murine models of visceral pain (Basso et al., 2018, 2016; Boué et al., 2014). However, the specific involvement of Treg in this process has not been investigated. Further research is required to determine the contribution of Treg-derived enkephalins in models of intestinal inflammatory pain. In neuroinflammatory settings, such as sciatic nerve constriction, the pain modulation by Treg could be contingent upon their immuno-suppressive function (Davoli-Ferreira et al., 2020; Duffy et al., 2019; Ledeboer et al., 2007; Shen et al., 2013), potentially mediated through the secretion of IL-10 or IL-35 (Davoli-Ferreira et al., 2020; Duffy et al., 2019). Contrary to this notion, our experiments rule out the possibility that hyperalgesia stemmed from a defect in the immuno-suppressive function of Treg. Instead, they strongly indicate a direct implication of enkephalins produced by Treg in nociception, revealing a novel non-immune intrinsic role of Treg in the endogenous regulation of basal somatic sensitivity.

Supporting information